---
title: |
  Exploratory Data Analysis \
  Conflict Data from ACLED
author: "Varvara Ilyina"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: no
    theme: lumen
    toc: yes
    toc_depth: 5
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---
<style>
div.answer {background-color:#f3f0ff; border-radius: 5px; padding: 20px;}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA)
```

### Requirements
What are some of the most important patterns in your data that you can observe through exploratory data analysis?
For instance, what does the label look like?
What do the simple bivariate relationships in the data look like?
Will this affect any of your modeling choices?

<hr>
<br>

### Setup
```{r, include = TRUE, echo = TRUE, results = 'hide'}

## load packages
pacman::p_load(reticulate, tidyverse, haven, readxl, writexl, stringr, ggplot2, sf, maps, viridis, modelsummary, gt, webshot2, knitr, kableExtra, janitor, corrplot, report)

## check working directory
getwd()

## set colors
#COL2 <- colorRampPalette(c("goldenrod3", "white", "purple1"))

## create theme for ggplot
angle_theme <- theme(plot.title = element_text(hjust = 0.5, size = 14),
                     axis.title.y = element_text(vjust = 1, size = 12),
                     axis.text.y = element_text(size = 10),
                     axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
                     plot.subtitle = element_text(size = 10),
                     plot.caption = element_text(),
                     legend.text = element_text(size = 11),
                     legend.key.size = unit(2, 'lines'),
                     legend.title = element_blank())

```
<br>

## Import / clean data
```{r, include = TRUE, echo = TRUE, results = 'hide'}

## load merged dataset
data_raw <- read.csv("../2_Dataset/d_Final_Dataset/Final_Dataset_Lag.csv", sep = ",")

## load final clean dataset
data_final_raw <- read.csv("../2_Dataset/d_Final_Dataset/Final_Dataset.csv")

## clean data
data <- data_final_raw %>% janitor::clean_names() %>%
  
  # rename zcta variable
  rename(zcta = zcta_conf) %>%
  
  # remove ", PA" from city names
  mutate(name_of_zip_code = str_extract(name_of_zip_code, "^.*(?=(,))"))
  

## load all ACLED data
data_acled_all <- read.csv("../2_Dataset/a_Input_Datasets/US2020_2023.csv", sep = ",")

```
<br>

## Choose important features
```{r, include = TRUE, echo = TRUE}

## select features marked as important for EDA
codebook <- read_excel("./codebook_clean.xlsx") %>%
  filter(important == 1) %>%
  pivot_wider(names_from = var_name,
              values_from = var_length) %>%
  select(-c(1:6)) %>%
  janitor::clean_names()

## vector with important vars
vars_keep <- names(codebook)

```
<br>








## Zip codes with most events
```{r, include = TRUE, echo = TRUE}

## calculate number of events per zip code
df <- data %>%
  select(zcta, month, event_count) %>%
  filter(event_count > 0) %>%
  group_by(zcta) %>%
  summarize(tot_events = sum(event_count)) %>%
  ungroup() %>%
  select(zcta, tot_events) %>%
  unique() %>%
  arrange(desc(tot_events))

## df with zip code names and pop
df_tmp <- data %>%
  select(zcta, admin2, name_of_zip_code, total_population_2022, population_per_square_land_mile_2022) %>%
  unique() %>%
  
  # remove ", PA" from city names
  mutate(name_of_zip_code = str_extract(name_of_zip_code, "^.*(?=(,))"))


## merge names onto zip code df
df_all <- left_join(df, df_tmp, by = "zcta") %>%
  filter(tot_events > 4)

```
<br>

```{r, include = TRUE, echo = TRUE}

caption <- "Only post office city names in Pennsylvania with more than 5 conflicts in 2022 included, \n where post office city names correspond with unique zip codes"

## plot event freq by zip
p <- ggplot(df_all, aes(reorder(name_of_zip_code, -tot_events), tot_events, fill = factor(name_of_zip_code))) +
  geom_bar(stat = "identity", position = "dodge", fill = "steelblue3") +
  labs(title = "Number of Conflicts per PA Zip Code in 2022",
       #subtitle = "Ordered in decreasing order",
       x = "",
       y = "Number of Conflicts",
       caption = caption) +
  scale_y_continuous(limits = c(0, max(df_all$tot_events + 20))) +
  #geom_text(aes(name_of_zip_code, tot_events, label = str_c(tot_events, " %")),
  geom_text(aes(name_of_zip_code, tot_events, label = tot_events),
            position = position_dodge(0.9), vjust = -0.5, size = 3) +
  angle_theme + theme(legend.position = "none")

print(p)

## save plot
ggsave("./figures/zips_events.png", plot = p, width = 10, height = 6, dpi = 300)

```
<br>

```{r, include = TRUE, echo = TRUE}

## set title
title <- "Number of Conflicts vs. Population Density \n in PA Counties with Highest Number of Conflicts in 2022"

## plot pop density vs. events
p <- ggplot(df_all, aes(x = tot_events, y = population_per_square_land_mile_2022, color = factor(name_of_zip_code))) +
  geom_point(size = 3) +
  labs(title = title,
       x = "Number of Conflict Events",
       y = "Population Density",
       caption = "Only includes counties with more than 5 conflicts in 2022") +
  geom_text(aes(label = name_of_zip_code), hjust = 1, vjust = -1, size = 3, color = "black") +
  angle_theme + 
  theme(legend.position = "none")

print(p)

## save plot
ggsave("4_Documentation/figures/scatter_pop_events_zips.png", plot = p, width = 10, height = 6, dpi = 300)


## plot pop density for zips with most events
#p <- ggplot(df_all, aes(reorder(admin2, -tot_events), population_per_square_land_mile_2022, fill = factor(admin2))) +
#  geom_bar(stat = "identity", position = "dodge", fill = "steelblue3") +
#  labs(title = "Number of Conflicts in 
#       Population in PA Counties with Highest Number of Conflicts in 2022",
#       #subtitle = "subtitle",
#       x = "",
#       y = "Population",
#       caption = "Only includes counties with more than 5 conflicts in 2022") +
#  scale_y_continuous(limits = c(0, max(df_all$population_per_square_land_mile_2022 + 20))) +
#  #geom_text(aes(admin2, tot_events, label = str_c(tot_events, " %")),
#  geom_text(aes(admin2, tot_events, label = tot_events),
#            position = position_dodge(0.9), vjust = -0.5, size = 3) +
#  angle_theme + theme(legend.position = "none")

#print(p)

```
<br>

## Months with most events
```{r, include = TRUE, echo = TRUE}

## create df with total events per month
df <- data %>%
  select(zcta, month, event_count) %>%
  filter(event_count > 0) %>%
  group_by(month) %>%
  summarize(tot_events = sum(event_count)) %>%
  ungroup()


## plot event freq by month
p <- ggplot(df, aes(month, tot_events, fill = factor(month))) +
  geom_bar(stat = "identity", position = "dodge", fill = "steelblue3") +
  labs(title = "Number of Conflicts per Month in PA in 2022",
       x = "",
       y = "Number of Conflicts") +
  scale_y_continuous(limits = c(0, max(df$tot_events + 20))) +
  geom_text(aes(month, tot_events, label = tot_events),
            position = position_dodge(0.9), vjust = -0.5, size = 4) +
  angle_theme + theme(legend.position = "none")

print(p)

## save plot
ggsave("4_Documentation/figures/months_events.png", plot = p, width = 10, height = 6, dpi = 300)

```
<br>

## Use lag data
```{r, include = TRUE, echo = TRUE}

## create df with lag variables
df_lag <- data %>%
  select(zcta, month, name_of_zip_code, event_count, neighbor_conflicts, event_count_prev_year, event_count_prev_month)

```
<br>


#### Event counts by month
```{r, include = TRUE, echo = TRUE}

## create df with month data
df <- df_lag %>%
  select(zcta, month, event_count, event_count_prev_month) %>%
  
  # create var if conflict in current and previous month
  group_by(zcta) %>%
  mutate(
    conflict_persists = ifelse(event_count > 0 & event_count_prev_month > 0, 1, 0)
  ) %>%
  ungroup() %>%
  filter(conflict_persists == 1) %>%
  select(-conflict_persists)


## vector with zips where conflict persists
problematic_zips <- unique(df$zcta)


## prepare df for plot
df_tmp <- df %>%
  filter(zcta == "19107") %>%
  pivot_longer(cols = c(event_count, event_count_prev_month), names_to = "variable", values_to = "value")


## plot
p <- ggplot(df_tmp, aes(month, value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  
  labs(title = "Number of Conflicts in Philadelphia (19107) per Month in PA in 2022",
       x = "",
       y = "Number of Conflicts") +
  scale_y_continuous(limits = c(0, max(df_tmp$value + 20))) +
  geom_text(aes(month, value, label = value),
            position = position_dodge(0.9), vjust = -0.5, size = 4) +
  angle_theme + theme(legend.position = "none")

print(p)


### this is stupid and obvious



# lol
ggplot(df_tmp, aes(x=month, y=value, fill=variable)) +
    geom_bar(stat='identity', position='dodge')




## save plot
#ggsave("4_Documentation/figures/months_events.png", plot = p, width = 10, height = 6, dpi = 300)

```
<br>










## Events across US states - table
```{r, include = TRUE, echo = TRUE}

df <- data_acled_all %>%
  filter(year == 2022) %>%
  select(event_id_cnty, event_date, admin1)

df %>%
  group_by(admin1) %>%
  rename(us_state = admin1) %>%
  mutate(tot_events_2022 = n()) %>%
  ungroup() %>%
  select(us_state, tot_events_2022) %>%
  unique() %>%
  arrange(desc(tot_events_2022)) %>%
  slice(1:10) %>%
  gt() %>%
  gtsave("./tables/events_by_state.png")

```
<br>

## Events across US states - map
```{r, include = TRUE, echo = TRUE}

## prepare acled data
df <- data_acled_all %>%
  filter(year == 2022) %>%
  select(event_id_cnty, event_date, admin1)

## calculate conflict events by US state
df_tmp <- df %>%
  group_by(admin1) %>%
  rename(us_state = admin1) %>%
  mutate(tot_events_2022 = n()) %>%
  ungroup() %>%
  select(us_state, tot_events_2022) %>%
  unique() %>%
  arrange(desc(tot_events_2022))

## get us states map data
us_states <- map_data("state")

## convert state names to lowercase to match the map data
df_tmp$us_state <- tolower(df_tmp$us_state)

## merge the event data with the map data
map_data <- us_states %>%
  left_join(df_tmp, by = c("region" = "us_state"))

## create heat map
p <- ggplot(data = map_data, aes(x = long, y = lat, group = group, fill = tot_events_2022)) +
  geom_polygon(color = "black") +
  scale_fill_viridis_c(option = "plasma", direction = -1, na.value = "white") +
  labs(title = "Number of Conflict Events by US State in 2022",
       fill = "Event Count",
       x = "",
       y = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

print(p)

## save plot
ggsave("4_Documentation/figures/us_heatmap.png", plot = p, width = 10, height = 6, dpi = 300)

```




